Constraints on Assembly Bias from Galaxy Clustering

### 2016-06-16 ms.tex
Edited section 2, adding many details and citing CorrFunc.

\author[Zentner et al.]
{Andrew R. Zentner$^{1}$, Andrew Hearin$^{2}$, Frank C. van den Bosch$^{3},$ \newauthor
Antonio Villareal$^{1}$, Johannes U. Lange$^{3}$, Kuan Wang$^{1}$, and P. Rogers Nelson$^{4}$\\ \\
$^1$Department of Physics and Astronomy \& Pittsburgh Particle Physics, Astrophysics, and Cosmology Center (PITT PACC),\\ University of Pittsburgh, Pittsburgh, PA 15260\\
$^2$Yale Center for Astronomy \& Astrophysics, Yale University, New Haven, CT\\
$^3$Department of Astronomy, Yale University, P.O. Box 208101, New Haven, CT\\
$^4$Paisley Park, Chanhassen, MN\\
}

@@

For more than a decade, halo occupation modeling has been used to
interpret large-scale structure measurements and exploit these
measurements to constrain galaxy formation models and cosmology
\citep[e.g.,][]{yang03,tinker05,zehavi05a,
  porciani06,vdBosch07,Zheng07,conroy_wechsler09,yang09b,zehavi_etal11,guo_etal11b,
  wake_etal11,yang11a,yang12,leauthaud_etal12,rod_puebla12,tinker_etal13,cacciato_etal13,
  more_etal13,guo_etal14,zu_mandelbaum15b}. The key assumptions
underlying halo occupation modeling are: (1) all galaxies reside in
dark matter halos that are biased tracers of the density field; and
(2) galaxies occupy halos as a function of halo mass $\mvir$ only. 
In this paper, we present the first analysis of large-scale clustering using 
a halo occupation model that violates the second assumption and permits 
galaxies to occupy halos as a function of multiple halo properties.

@@

To generate predictions for galaxy clustering, we directly populate dark matter halos with mock galaxies using {\tt Halotools}.
We explore halo occupation distribution models (HOD) models in this work
\citep[e.g.][]{seljak00,ma_fry00,scoccimarro01a,berlind02}, though other
techniques that can be used to interpret such data, such as the conditional luminosity function
\citep[CLF, e.g.,][]{yang03,vdBosch13}, exist. In this subsection, we review the ``standard" HOD
model used in the present work, which assumes that there is no galaxy assembly bias. We refer to
such a model as ``standard" because all HOD analyses of galaxy clustering to date have assumed no
galaxy assembly bias. In the following subsection, we describe
the Decorated HOD model described in \citet{hearin_etal16}. In both cases, we will only
review the salient features of our methodology briefly;
interested readers can always refer to {\tt http://halotools.readthedocs.io} and
\citet{hearin_etal16} for further details.

@@

All of our analyses are based on the Bolshioi-Planck (BolshioP) simulation \citep{riebe_etal11}.
BolshoiP was run with cosmological parameters based on \citet{planck13},
$\Omega_{\Lambda} = 0.693, \Omega_{\rm m} = 1 - \Omega_{\Lambda} = 0.307, \Omega_{\rm b} = 0.048,
h = 0.7, {\rm n_s}=0.96$ and $\sigma_8 = 0.82,$ within a cubic box
$250$ ${\rm Mpc}/h$ on a side, requiring a particle mass of $m_{\rm p} = 1.35\times10^{8}M_{\odot}/h.$
Further information about the BolshoiP simulation is available at \url{https://www.cosmosim.org}.

We use publicly available\footnote{\tt http://www.slac.stanford.edu/$\sim$behroozi/BPlanck\_Hlists} dark matter
halo catalogs based on the {\tt ROCKSTAR} halo-finder \citep{behroozi_rockstar11} and {\tt CONSISTENT TREES} algorithm \citep{behroozi_trees13}. In particular, we use the {\tt halotools\_alpha\_version2} version of the $z=0$ snapshot
of the {\tt `bolplanck'} catalog included with {\tt Halotools.} Halos in these catalogs are based on the virial radius
density contrast given in \citet{bryan_norman98}. When populating this catalog with mock galaxies, we only use
present-day host halos with a value of $M_{\rm peak}$ that exceeds $300$ particles.

@@

In standard HOD models, central galaxies and satellite galaxies are treated separately, so
the model is specified by two probability distributions, one for each type of galaxy.
The galaxy-halo connection is specified in terms of $P(\ncen|\mvir)$ and $P(\nsat|\mvir),$
the probability that a halo of mass $\mvir$ hosts $\ncen$ central and $\nsat$ satellite galaxies,
respectively. $P(\ncen|\mvir)$ is typically a nearest-integer distribution, as a host halo has only
either zero or one central galaxy. Consequently, the occupation statistics of central galaxies are
specified by the first moment of $P(\ncen|\mvir)$, which we model as
%
\begin{equation}
\label{eq:ncen}
\langle N_{\mathrm{cen}} \rangle(\mvir) =
        \frac{1}{2}\left( 1 +
        \mathrm{erf}\left( \frac{\log (\mvir) -
        \log (M_{\rm min})}{\sigma_{\log M}} \right) \right)
\end{equation}
%
For every host halo in the catalog we draw a random number from a uniform distribution $\mathcal{U}(0, 1);$ for a host halo of present-day virial mass $\mvir,$ a central galaxy is assigned to the halo if the associated random number is less than $\langle N_{\mathrm{cen}} \rangle(\mvir);$ halos with random values exceeding  $\langle N_{\mathrm{cen}} \rangle(\mvir)$ are left devoid of centrals.
The parameter $\log (M_{\rm min})$ specifies the halo mass at which the halo has a 50\% probability of hosting a
central galaxy, while the parameter $\sigma_{\log M}$ specifies the rate at which $\langle N_{\mathrm{cen}} \rangle (\mvir)$
transitions from zero to unity, with smaller values of $\sigma_{\log M}$ corresponding to a more rapid transition.

We model the distribution $P(\nsat|\mvir)$ as a Poisson distribution with first moment given by a power law,
%
\begin{equation}
\label{eq:nsat}
\langle N_{\rm sat}\rangle(\mvir) = \left( \frac{\mvir - M_0}{M_1} \right)^{\alpha}.
\end{equation}
%
The parameter $M_0$ allows the power-law to be truncated more rapidly at low masses and we
set $\langle N_{\rm sat}\rangle (\mvir) = 0$ for $\mvir < M_0$.

The five parameters of HOD models that are varied in standard analyses are $\log (M_{\rm min})$,
$\sigma_{\log M}$, $\alpha$, $\log (M_1)$, and $\log (M_0)$, though, as we show below, central galaxies usually
outnumber satellite galaxies by a factor of several, so $\log (M_{\rm min})$ and $\sigma_{\log M}$ usually
vary along a narrow degeneracy that fixes the total galaxy number density to the observed value. There are
many particular choices that can be made for the functional forms of $\langle N_{\rm cen}\rangle (\mvir)$ and
$\langle N_{\rm sat}\rangle (\mvir)$. We have made choices that mimic the standard SDSS DR7 analysis of
\citet{zehavi_etal11}, to expedite comparisons with their results.

@@

After populating a halo catalog with mock galaxies, we calculate the
comoving number density of our mock galaxy sample as $n_{\rm g} =
N_{\rm gal} / L_{\rm box}^3$, where $N_{\rm gal}$ is the total number of
galaxies in the mock sample. We apply the distant-observer
approximation and use the simulation z-axis as the line-of-sight
direction, and the distance between points in the $xy$-plane to define
the projected distance $\rp.$ We place mock galaxies into
redshift-space by replacing each galaxy's $z$-coordinate with
$z_{\rm RS} = z + V_{\rm z}/H_0.$ We count pairs of points in each of
our $\rp$ bins, rejecting pairs with $z_{\rm RS}-$ distance exceeding
$\pi_{\rm max}=60$ ${\rm Mpc/h}$, which is the same projection
depth chosen by \citet{zehavi_etal11}. We employ the \citet{landyszalay93}
estimator to turn pair counts into a $\wprp$ prediction, as described
in detail in {\tt http://halotools.readthedocs.io}.

@@

In addition to the standard occupation statistics described in the
previous section, in this paper we also use the decorated HOD
formalism to connect galaxies to dark matter halos in a manner that
has simultaneous dependence on both $\mvir$ and {\em halo concentration}.
Briefly, we use Equations \ref{eq:ncen} and \ref{eq:nsat} as our
``baseline" first occupation moments.  At fixed $\mvir,$ halos are
divided into one of two categories, those of high- and
low-concentration, depending on whether the concentration of the halo
places it above or below the rank-order percentile $f_{\rm split},$
which we keep fixed to $f_{\rm split}=0.5$ throughout the paper for
simplicity. High-concentration halos have a different first
occupation moment relative to low-concentration halos of the same
mass, $\langle\ngal|\mvir, c_{\rm high}\rangle \neq
\langle\ngal|\mvir, c_{\rm low}\rangle.$ The difference between the
first moment of high- and low-concentration halos is modulated by
$\abias,$ the novel parameter of the decorated HOD governing assembly
bias. Values of $\abias=\pm1$ correspond to the maximum strength of
assembly bias allowable by the constraint that the model preserves the
marginalized first moment, $\langle\ngal|\mvir\rangle;$ thus
regardless of the value of $\abias,$ in the decorated HOD the
marginalized first moment of centrals and satellites are {\em
  unchanged from the baseline value defined by Equations \ref{eq:ncen}
  and \ref{eq:nsat}}. In other words, {\em decorated HOD models all have the same HODs,
when averaged over all halos at fixed mass, as standard HOD
models}. The only change in Decorated HOD models is whether or not an
additional property also modulates halo occupation at fixed halo
mass. A value of $\abias = 0$ indicates no galaxy assembly bias
whatsoever. We refer the reader to \citet{hearin_etal16} for further
details about the decorated HOD.

>>>>

\author[Zentner et al.]
{Andrew R. Zentner$^{1}$, Andrew Hearin$^{2}$, Frank C. van den Bosch$^{3},$ \newauthor
Johannes U. Lange$^{3}$, and P. Rogers Nelson$^{4}$\\ \\
$^1$Department of Physics and Astronomy \& Pittsburgh Particle Physics, Astrophysics, and Cosmology Center (PITT PACC),\\ University of Pittsburgh, Pittsburgh, PA 15260\\
$^2$Yale Center for Astronomy \& Astrophysics, Yale University, New Haven, CT\\
$^3$Department of Astronomy, Yale University, P.O. Box 208101, New Haven, CT\\
$^4$Paisley Park, Chanhassen, MN\\
}

@@

For more than a decade, halo occupation modeling has been used to
interpret large-scale structure measurements and exploit these
measurements to constrain galaxy formation models and cosmology
\citep[e.g.,][]{yang03,tinker05,zehavi05a,
  porciani06,vdBosch07,Zheng07,conroy_wechsler09,yang09b,zehavi_etal11,guo_etal11b,
  wake_etal11,yang11a,yang12,leauthaud_etal12,rod_puebla12,tinker_etal13,cacciato_etal13,
  more_etal13,guo_etal14,zu_mandelbaum15b}. The key assumptions
underlying halo occupation modeling are: (1) all galaxies reside in
dark matter halos that are biased tracers of the density field; and
(2) galaxies occupy halos as a function of halo mass $\mvir$ only. 
In this paper, we present the first analysis of galaxy clustering using 
a halo occupation model that violates the second of these assumptions 
and permits galaxies to occupy halos as a function of multiple halo properties.

@@

To generate predictions for galaxy clustering, 
we populate dark matter halos with mock galaxies using the publicly-available, 
open source, {\tt Halotools} software \citep{halotools}.
We explore halo occupation distribution (HOD) models in this work
\citep[e.g.][]{seljak00,ma_fry00,scoccimarro01a,berlind02}, though other
techniques that can be used to interpret such data, such as the conditional luminosity function
\citep[CLF, e.g.,][]{yang03,vdBosch13}, exist. In this subsection, we review the ``standard" HOD
model used in the present work, which assumes that there is no galaxy assembly bias. We refer to
such a model as ``standard" because all HOD analyses of galaxy clustering to date have assumed no
galaxy assembly bias. In the following subsection, we describe
the Decorated HOD model described in \citet{hearin_etal16}. In both cases, we will only
review the salient features of our methodology briefly;
interested readers can always refer to {\tt http://halotools.readthedocs.io} and
\citet{hearin_etal16} for further details.

@@

All of our analyses are based on the Bolshioi-Planck (BolshioP) simulation \citep{riebe_etal11}.
BolshoiP was run with cosmological parameters based on \citet{planck13}:
$\Omega_{\Lambda} = 0.693; \Omega_{\rm m} = 1 - \Omega_{\Lambda} = 0.307; \Omega_{\rm b} = 0.048;
h = 0.7; {\rm n_s}=0.96$; and $\sigma_8 = 0.82$. BolshoiP simulated the formation of structure within a cubic 
box $250$ ${\rm Mpc}/h$ on a side, requiring a particle mass of $m_{\rm p} = 1.35\times10^{8}M_{\odot}/h.$
Further information about the BolshoiP simulation is available at \url{https://www.cosmosim.org}.

We use publicly available\footnote{\tt http://www.slac.stanford.edu/$\sim$behroozi/BPlanck\_Hlists} dark matter
halo catalogs based on the {\tt ROCKSTAR} halo-finder \citep{behroozi_rockstar11} and {\tt CONSISTENT TREES} algorithm \citep{behroozi_trees13}. In particular, we use the {\tt halotools\_alpha\_version2} version of the $z=0$ snapshot
of the {\tt `bolplanck'} catalog included with {\tt Halotools.} Halos in these catalogs are based on the virial radius
density contrast given in \citet{bryan_norman98}. When populating this catalog with mock galaxies, we only use
present-day host halos with a value of $M_{\rm peak}$ that exceeds $300$ particles. The minimum 
peak halo mass considered in our analysis has $M_{\rm peak} \ge 4.05 \times 10^5\, h^{-1}\mathrm{M}_{\odot}$ 
and this prevents us from analyzing samples with $M_r > -19$.

@@

In standard HOD models, central galaxies and satellite galaxies are treated separately, so
the model is specified by two probability distributions, one for each type of galaxy.
The galaxy-halo connection is specified in terms of $P(\ncen|\mvir)$ and $P(\nsat|\mvir),$
the probability that a halo of mass $\mvir$ hosts $\ncen$ central and $\nsat$ satellite galaxies,
respectively. $P(\ncen|\mvir)$ is typically a nearest-integer distribution, as a host halo has only
either zero or one central galaxy. Consequently, the occupation statistics of central galaxies are
specified by the first moment of $P(\ncen|\mvir)$, which we model as
%
\begin{equation}
\label{eq:ncen}
\langle N_{\mathrm{cen}} \vert \mvir \rangle =
        \frac{1}{2}\left( 1 +
        \mathrm{erf}\left( \frac{\log (\mvir) -
        \log (M_{\rm min})}{\sigma_{\log M}} \right) \right)
\end{equation}
%
For every host halo in the catalog we draw a random number from a uniform distribution $\mathcal{U}(0, 1);$ for a host halo of present-day virial mass $\mvir,$ a central galaxy is assigned to the halo if the associated random number is less than $\langle N_{\mathrm{cen}} \vert \mvir \rangle;$ halos with random values exceeding  $\langle N_{\mathrm{cen}} \vert \mvir \rangle$ are left devoid of centrals.
The parameter $\log (M_{\rm min})$ specifies the halo mass at which the halo has a 50\% probability of hosting a
central galaxy, while the parameter $\sigma_{\log M}$ specifies the rate at which $\langle N_{\mathrm{cen}} \vert \mvir \rangle$
transitions from zero to unity, with smaller values of $\sigma_{\log M}$ corresponding to a more rapid transition.

We model the distribution $P(\nsat|\mvir)$ as a Poisson distribution with first moment given by a power law,
%
\begin{equation}
\label{eq:nsat}
\langle N_{\rm sat}\vert \mvir \rangle = \left( \frac{\mvir - M_0}{M_1} \right)^{\alpha}.
\end{equation}
%
The parameter $M_0$ allows the power-law to be truncated more rapidly at low masses and we
set $\langle N_{\rm sat}\vert \mvir \rangle = 0$ for $\mvir < M_0$.

The five parameters of HOD models that are varied in standard analyses are $\log (M_{\rm min})$,
$\sigma_{\log M}$, $\alpha$, $\log (M_1)$, and $\log (M_0)$, though, as we show below, central galaxies usually
outnumber satellite galaxies by a factor of several, so $\log (M_{\rm min})$ and $\sigma_{\log M}$ usually
vary along a narrow degeneracy that fixes the total galaxy number density to the observed value. There are
many particular choices that can be made for the functional forms of $\langle N_{\rm cen}\vert \mvir \rangle$ and
$\langle N_{\rm sat}\vert \mvir \rangle$. We have made choices that mimic the standard SDSS DR7 analysis of
\citet{zehavi_etal11}, to expedite comparisons with their results. However, our choice does deviate from 
the choice of \citet{zehavi_etal11} in one respect. The mean satellite occupation of \citet{zehavi_etal11} 
is that of Eq.~(\ref{eq:nsat}) multiplied by an overall factor of $\langle N_{\rm cen} \vert \mvir \rangle$. 
We have chosen not to use this as our default because it can introduce difficulties in some analyses 
\citep[see the discussion of blue galaxy samples in][]{zentner_etal14} and because, as we have verified 
explicitly, the extra factor introduces only small quantitative changes to our results and no qualitative 
changes.

@@

After populating a halo catalog with mock galaxies, we calculate the
comoving number density of our mock galaxy sample as $n_{\rm g} =
N_{\rm gal} / L_{\rm box}^3$, where $N_{\rm gal}$ is the total number of
galaxies in the mock sample. We apply the distant-observer
approximation and use the simulation z-axis as the line-of-sight
direction, and the distance between points in the $xy$-plane to define
the projected distance $\rp.$ We place mock galaxies into
redshift-space by replacing each galaxy's $z$-coordinate with
$z_{\rm RS} = z + V_{\rm z}/H_0.$ We count pairs of points in each of
our $\rp$ bins, rejecting pairs with line-of-sight distances in redshift space, $\Delta z_{\rm RS}$, 
exceeding $\pi_{\rm max}=60$ ${\rm Mpc/h}$, which is the same projection
depth chosen by \citet{zehavi_etal11}. Having populated mocks, 
we perform the computation of the projected two-point galaxy correlation 
function, $w_{\rm p}(r_{\rm p})$ using the publicly-available {\tt CorrFunc} 
package \citep{corrfunc} which has been extensively optimized for computational speed.

@@

In addition to the standard occupation statistics described in the
previous section, in this paper we also use the decorated HOD
formalism to connect galaxies to dark matter halos in a manner that
has simultaneous dependence on both $\mvir$ and {\em halo concentration}.
Briefly, we use Equations \ref{eq:ncen} and \ref{eq:nsat} as our
``baseline" first occupation moments.  At fixed $\mvir,$ halos are
divided into one of two categories, those of high- and
low-concentration, depending on whether the concentration of the halo
places it above or below the rank-order percentile $f_{\rm split},$
which we keep fixed to $f_{\rm split}=0.5$ throughout the paper for
simplicity. High-concentration halos have a different first
occupation moment relative to low-concentration halos of the same
mass, $\langle\ngal|\mvir, c_{\rm high}\rangle \neq
\langle\ngal|\mvir, c_{\rm low}\rangle.$ The difference between the
first moment of high- and low-concentration halos is modulated by
$\abias,$ the novel parameter of the decorated HOD governing assembly
bias. Values of $\abias=\pm1$ correspond to the maximum strength of
assembly bias allowable by the constraint that the model preserves the
marginalized first moment, $\langle\ngal|\mvir\rangle;$ thus
regardless of the value of $\abias,$ in the decorated HOD the
marginalized first moment of centrals and satellites are {\em
  unchanged from the baseline value defined by Equations \ref{eq:ncen}
  and \ref{eq:nsat}}. In other words, {\em decorated HOD models all have the same HODs,
when averaged over all halos at fixed mass, as standard HOD
models with the same five standard HOD parameters}. 
The only change in Decorated HOD models is whether or not an
additional property also modulates halo occupation at fixed halo
mass. A value of $\abias = 0$ indicates no galaxy assembly bias
whatsoever. We refer the reader to \citet{hearin_etal16} for further
details about the decorated HOD.
